\section{Introduction}\label{sec:introduction}

\subsection*{What is Knowledge Representation?}
\defin{kr}

The core problem with KR is that the real world with its complexity, ambiguity and dynamism can never be fully represented in a computer-based system.
Nevertheless, intelligent systems must be able to function effectively using partial, incomplete and sometimes contradictory information.

\subsection*{Why is Knowledge Representation needed?}

\subsubsection*{The problem of data meaninglessness}
To understand the need for KR, let’s consider a practical scenario:
\examp{
    A student database.

    A hypothetical database contains the following entries:

    \begin{table}[H]
        \centering
        \begin{tabular}{lllll}
            \textbf{Name} & \textbf{Major} & \textbf{CGPA} & \textbf{Research}   & \ldots \\
            Alice Smith   & CS             & 3.8           & Genomics            & \ldots \\
            Bob John      & Biology        & 3.8           & Oncology            & \ldots \\
            Carol Li      & Zoology        & 3.7           & Animal colonisation & \ldots
        \end{tabular}
        \label{tab:first-example}
    \end{table}

    Without knowledge representation, this data is disjointed.
    The entries exist in isolation across thousands of files.
    A database query system can only perform exact keyword matches:

    \begin{itemize}
        \item Query 1: ``Show me all AI researchers'' → No results (not explicitly mentioned in the data)
        \item Query 2: ``Which students could work on this patent?'' → No answer possible
        \item Query 3: ``What are the emerging trends?'' → No analysis possible
    \end{itemize}
}

Data is meaningless without context and relationships.
It does not allow for intelligent reasoning or deeper analysis.

\subsection*{What knowledge representation makes possible}

With knowledge representation, we can structure the data and transfer isolated records into interconnected,
queryable knowledge using:
\begin{itemize}
    \item \glspl{ontology}: Explicitly specify which concepts exist (student, researcher, field of research)
    and how they are categorised.
    \item Knowledge graphs: Link entities together: Alice → researcher → AI
\end{itemize}

This allows for semantic searches, automatic relationship discovery and reasoning across all connection.

\examp{
    \begin{itemize}
        \item Semantic search
        \begin{itemize}
            \item ``Find students researching machine learning?''
            \item ``Show me all AI researchers'': The system can use transitive relationships to find relevant students.
            \item ``Which students are a good fit for this patent?'': The system can match skills and areas of interest.
        \end{itemize}
        \item Automatic relationship discovery
        \begin{itemize}
            \item ``What are emerging research trends?''
            \item Recognising patterns across all data
            \item Identifying emerging trends through interconnected information
        \end{itemize}
    \end{itemize}
}

\begin{table}[H]
    \begin{tabular}{p{5cm}p{5cm}p{5cm}}
        \Gls{data}      & \Gls{information}                                 & \Gls{knowledge}                   \\
        Raw, unprocessed facts & Structured and contextualised data & Information combined with experience and
        reasoning rules \\
        e.g.\ CGPA: 3.8 & e.g.\ Alice has a CGPA of 3.8 in Computer Science & e.g.\ Alice could work on this AI
        patent because she has a strong foundation in computer science and is interested in genomics
    \end{tabular}
    \caption{Data vs. Information vs. Knowledge }\label{tab:data-information-knowledge}
\end{table}

KR enables machines to make informed decisions with fewer rules.

\subsection*{KR Perspective}

\subsection*{Representation Schemes}

\subsection*{Challenges in Knowledge Representation}

\paragraph{Traditional ML} systems follow a black-box approach which means the decision-making process is not transparent and the reasoning behind predictions is not accessible.

\begin{itemize}
    \item Generalisation: Requires continuous retraining with new data
    \item Explainability: Difficult or impossible to explain decisions
    \item Data flexibility: Highly dependent on large amounts of training data
\end{itemize}

\paragraph{Knowledge-Based AI} utilises structured, \gls{explicit knowledge}.

\defin{explicit knowledge}

\begin{itemize}
    \item Generalization: explicit knowledge transfers to new problems without retraining
    from scratch\cite{hogan2021}
    \item Explainability: encoded knowledge enables transparent reasoning and human-understandable decisions\cite{cao2024}
    \item Data flexibility: Modifiable
\end{itemize}

\subsection*{Representation Schemes}
\defin{representation scheme}

\examp{
    \textbf{Pac-Man ghost AI} has a knowledge-based approach: We code ghost behaviour as explicit rules.
    \begin{enumerate}
        \item IF player\_nearby THEN chase
        \item IF frightened THEN flee
        \item ELSE patrol
    \end{enumerate}
    Decisions are transparent, explainable, and modifiable without recomputing. (Ogland 2016)
}

\subsection*{Challenges in Knowledge Representation}
Knowledge representation involves designing formal structures that allow computers to store, interpret, and reason
about information and every design decision has advantages and caveats.

\paragraph{Choosing the representation} is one of the design challenges.
\Glspl{semantic network} are intuitive and visually understandable but have limited inference. \Gls{first-order logic} on the other hand is
expressive, but has computationally costly reasoning.

\paragraph{Desired properties} can conflict.
Systems should ideally be expressive enough to capture rich real-world knowledge while remaining efficient enough to reason in practical time.
Tradeoffs are between expressiveness and tractability, and Human readability and machine efficiency.

\paragraph{Application-driven constraints} arise from, different domains requiring different forms of representation.
E.g.\ In natural language processing (NLP), systems must handle ambiguity and context.
In medical diagnosis, representations must manage uncertainty, causal relationships, and strict correctness
requirements.
In question answering, systems need fast retrieval combined with reasoning over large amounts of structured and unstructured data

\paragraph{The core idea of Knowledge Representation} is not just a data structure.
It plays multiple roles
simultaneously and those roles can conflict with each other: 
\begin{itemize}
    \item Surrogate for the world: Acts as a substitute for real world objects and relationships
    \item Ontological commitments: Defines what exists and how things are categorized
    \item Theory of reasoning: Provides foundation for logical inference and deduction
    \item Medium for efficient computation: Enables performant computational processing
    \item Medium of Human expression: Allows humans to express knowledge in understandable ways
\end{itemize}

\section{Taxonomy of Representation Schemes}\label{sec:taxonomy-of-representation-schemes}~\cite{folien1}


\section{Distinguishing Features of KR Systems}\label{sec:distinguishing-features-of-kr-systems}

Defis:
Frame, Inference,

kinds of reasoning
inference types (23+) David, Shrobe and Szolovist

Wie man mit Incompleteness umgeht


\section{Applications of Knowledge Representation}\label{sec:applications-of-knowledge-representation}

Enterprise Knowledge Management


\section{Future Directions}\label{sec:future-directions}

Multi-hop reasoning
Unified framework

knowledge Aggregation

def attention weight


\section{Course structure}\label{sec:course-structure}

foundation of KR

Knowledge graphs

KR in mL

grading:

50\% of the exercise points (full portfolio) (Abgabe in 3er Gruppen(?))
present one exercise in tut
present one paper in seminar 40--45 min (3er gruppen)
final report 5-.10 pages (50\% marks on final report)


Es wird ein Moodle geben


\section{Formalities}\label{sec:formalities}