%! Author = lili
%! Date = 4/30/26

%bis ende section 1 ging das 6--10

\section{The Maximal Matches Distance}\label{sec:the-maximal-matches-distance}
Another notion of distance that admits large block movements is the \gls{mmdis}.
It is based on the idea that we can cover a sequence $x$ with substrings of another
sequence (and vice versa) that are interspersed with single characters.
\paragraph{Partitioning a sequence.} Consider a sequence $x = z_0 b_1 z_1 b_2 \dots z_{l-1} b_l z_l$, with $|x| = m$,
such that each $z_i$ is a (possibly empty) string (for $0 \leq i \leq l$) and each $b_i$ is a single
character (for $1 \leq i \leq l$). The tuple 
$P = (z_0 , \langle b_1 \rangle , z_1 , \langle b_2 \rangle , \dots, z_{l-1} , \langle b_l \rangle , z_l )$ is a partition
of $x$ with respect to another sequence $y$ if each $z_i$ is also a substring of $y$.
The size of
the partition $P$ , denoted by $|P|$, is the number of its separating single characters, that is, $|P| = l$.
Note that, if $x$ is a substring of $y$, there exists the shortest partition $P_0 = (x)$, of
size 0.
In the other extreme, the longest partition $P_m = (\epsilon, \langle x[1]\rangle , \epsilon, \langle x[2]\rangle , \dots , \langle x[m]\rangle , \epsilon)$
of size $m$ always exists, even if $y = \epsilon$.

\begin{exa}
    Let $x = CTAATGCT$ and $y = ATCTA$.
    The following sequences are partitions of
    $x$ with respect to $y$: $P = (CTA, \langle A\rangle , T, \langle G\rangle , CT)$ with $P | = 2, P ' = (CTA, \langle A\rangle , T, \langle G\rangle , C, \langle T\rangle , \epsilon)$ with
    $|P ' | = 3$ and $P '' = (CT, \langle A\rangle , AT, \langle G\rangle , CT)$ with $|P '' | = 2$.
\end{exa}

\deffin{mmdis}

\subsection{Efficient Computation}\label{subsec:efficient-computation}
The next question is how to find a minimum size partition of $x$ with respect to $y$ among
all the partitions.
Fortunately, this turns out to be easy: A greedy strategy does the
job.
We start covering $x$ at the first position by a match of maximal length k such that
$z_0 \coloneqq x[1 \dots k]$ is a substring of $y$, but $z_0 b_1 = x[1 \dots k + 1]$ is not a substring of $y$.
We apply the same strategy to the remainder of $x$, $x[k + 2 \dots m]$.

\deffin{ltrpartition}

\deffin{rtlpartition}

\begin{theo} \label{thm:theorem-1-3}
    The \gls{ltrpartition} is a partition of minimal length, i.e.,
    \begin{center}
        $|\glssymbol{ltrpartition} (x, y)| = \min\{|P | : P $is a partition of $x$ with respect to $y\} = \mmsym(x||y)$.
    \end{center}
    The same is true for the \gls{rtlpartition}, which means:
    \[
        |\glssymbol{rtlpartition} (x, y)| = |\glssymbol{ltrpartition} (x, y)| = \mmsym (x||y).
    \]
    \paragraph{Proof.} Skript S. 7 % TODO
\end{theo}
It is important to discuss how (and how fast) we can compute the size of a \deffin{ltrpartition}.
In fact, because the longest common substring $z$ that starts at a given position
$i$ in $x$ and occurs somewhere in $y$ can easily be found in $O(|z|)$ time using the suffix tree
of $y$, we obtain the following result.

\begin{theo} \label{thm:theorem-1-4}
    The \gls{mmdis} $\mmsym(x||y)$ to a sequence $x$ from another sequence
    $y$ can be computed in $ O(|x| + |y|)$ time.
    \paragraph{Proof.} Skript S. 8 % TODO
\end{theo}

\subsection{Relation to the Edit Distance}\label{subsec:relation-to-the-edit-distance}

Like the \gls{qgdis}, also the \gls{mmdis} can be used to compute a
lower bound of the \gls{edis}.

\begin{theo} \label{thm:theorem-1-5}
    Let $d(x, y)$ be the unit cost \gls{edis} between sequences $x$ and $y$.
    Then $\mmsym(x||y) \leq d(x, y)$.
    \paragraph{Proof.} Skript S. 8 % TODO
\end{theo}

Again, this bound permits us to speed up the computation of the \gls{edis}, as will be
shown in Section~\ref{subsec:maximal-matches-metric}.

\subsection{Maximal Matches Metric}\label{subsec:maximal-matches-metric}
A final observation is that $\mmsym$ is obviously not symmetric: $\mmsym(x||y)  = 0$ is equivalent to
$x$ being a substring of $y$.
Therefore, the \gls{mmdis} is not a metric and the name distance is misleading.
But it is possible to use $\mmsym$ for designing an appropriate
symmetric function that satisfies the triangular inequality and is a metric.

\begin{theo} \label{thm:theorem-1-6}
    Given sequences $x$ and $y$ from $\Sigma^*$, let the function $d_{||} (x, y)$ defined as
    $d_{||} (x, y) = \log(\mmsym(x||y)+1) + \log(\mmsym(y||x)+1)$.
    Then $d_{||} (x, y)$ is a metric on $\Sigma^*$.
    \paragraph{Proof.} Skript S. 8 % TODO
\end{theo}

\section{Filtration for Edit Distance}\label{sec:filtration-for-edit-distance}
Coming back to the \gls{edis}, while – given the fact that the search space (consisting
of all pairwise alignments of $x$ and $y$) is of exponential size – the quadratic time dynamic
programming solution can be considered very efficient, it can also be considered quite
slow in practice when the sequences are long.
Moreover, the quadratic space requirements
(unless a space-saving technique is used that we will discuss in Chapter 6) can cause even
larger problems. % TODO chapter ref

Now, remember that in many applications the goal of sequence comparison is not the
construction of an accurate alignment.
Instead, for example in a database search context,
one is mostly interested in identifying sequences similar to a given query.
More formally:

\begin{prob}
    Given a database $X = {x_1 , x_2 , \dots , x_k }$ of $k$ sequences $x_i \in \Sigma^*$,
    a query sequence $y \in \Sigma^*$ and a distance threshold $t \geq 0$, find all
    sequences $x \in X$ such that $d(x, y) \leq t$.
\end{prob}

Clearly, a straightforward way to solve this problem is to compare $y$ to each sequence
$x \in X$, one after the other.
Whenever in such a comparison the distance $d(x, y)$ is smaller
or equal to the threshold $t$, then the sequence $x$ is reported, otherwise it is discarded.
This strategy clearly takes $O(k \cdot mn)$ time, where $m$ is an upper bound for the database
sequence length and $n = |y|$.

However, since usually $t$ is small and many sequences in $x$ will be quite dissimilar from
$y$,it may be possible to quickly screen the database and separate interesting candidates
from irrelevant sequences, faster than computing the \gls{edis} for all of them.
If such
a screening procedure guarantees that all relevant sequences $x \in X$ with $d(x, y) \leq t$ are
among the remaining candidates, it is called a filter.
Generally, we thus have a two-step procedure.
First, in a fast filtration step, candidates
are identified that have a chance to be hits.
Then, in a verification step these candidates
are checked with the (usually slower) exact method whether they really qualify as hits.
Obviously, filtration makes sense only if the procedure employed in the filtration step has
a lower time complexity than the procedure used in the verification step.
Indeed, if one has multiple filters, one can first apply all of them, and then employ the
verification only to those elements $x \in X$ that pass all the filters.
In the previous sections we have seen two fast sequence comparison methods that can be
used as filters for the unit cost \gls{edis}.
Both of them require only time proportional
to the sum of the lengths of the two sequences (“linear time”) and provide lower bounds
for the \gls{edis}.

\paragraph{Filtration with the \gls{qgdis}.} The property outlined in Theorem~\ref{thm:theorem-1-2} can be used
as a filter criterion for the \gls{edis}: For a given query sequence $y$ and \gls{edis}
threshold t, and a given value of q, a database sequence $x \in X$ can be discarded whenever
$\dqu (x, y)/(2q) > t$.
In case of a global similarity of the sequences and the differences spread
out evenly, this filter will work quite well.
If the difference between the two sequences,
however, is by block movements, then the \gls{edis} is usually very high, although the
\gls{qgdis} may still be low and produce a false positive candidate when used as a
filter.


\paragraph{Filtration with the \gls{mmdis}.} As shown in Theorem~\ref{thm:theorem-1-5}, the maximal
matches distance also gives us a bound for the unit cost \gls{edis}.
In fact, since the
\gls{edis} is symmetric, the filtration can even be used twice, once with $d(x||y)$ and
once with $d(y||x)$.
This results in the following filter: For a given query sequence $y$ and edit
distance threshold $t$, a database sequence $x \in X$ can be discarded whenever $\mmsym(x||y) >t$ or
$\mmsym(y||x) > t$.

Note that the distinguishing feature of a filter is that it never discards a hit (has no false
negatives).
On the other hand, one can also develop heuristics that approximate the edit
distance and are efficiently computable.
Good heuristics (like BLAST) will be close to the
true result with high probability; hence the error made by using them is small most of the
time.
There is no guarantee, though, that a true hit will not be missed.