%! Author = lili
%! Date = 4/30/26


A note on terminology: In Section~\ref{sec:the-q-gram-distance} we were speaking of q-grams when referring to
short (sub)strings of a certain length q.
This notation has originally been introduced in
the computer science literature and is there in use until today.
An alternative, equivalent
notion that one finds in most of the biological\footnote{The true origin is probably in biochemistry, where notions like monomer or polymer refer to molecular
chains of certain lengths.} and bioinformatics literature, is that of a
k-mer.
Here k refers to the (sub)string length.
To be more consistent with the literature,
in this chapter we will adopt that notation.

\section{Definition and Basic Properties}\label{sec:definition-and-basic-properties}
Remember the following definition, where $s_{k,i} \coloneqq s[i \dots i+k-1]$ denotes the k-mer starting
at index $i$, $1 \leq i \leq |s| -k +1$:
\defin{dbsg}
% V = alle k-mere von s

\defin{dimension}

\begin{notte} \textbf{To \autoref{def:dbsg}}
    If the same (k + 1)-mer occurs multiple times in s, say m times, we have m
    parallel edges occurring in the multigraph DBG(s, k).
\end{notte}

% \frage{} Todo Wie viele k-mere hat eine Sequenz der Länge n sollte man wissen: n-k+1

\komm{Für jeden DB Graph gibt es immer einen Eulerpfad}

\section{Words with the same (k + 1)-mer Profile}\label{sec:words-with-the-same-k-1-mer-profile}
Motivated by the fact that the q-gram distance does not satisfy the identity property of
a metric (see Section \ref{sec:the-q-gram-distance}) because there can be distinct sequences with the same q-gram
profile, we want to determine whether for a given sequence s and integer k there are other
sequences that are distinct but have the same (k + 1)-mer profile as s. And, if there are,
how many? \\ \newline
For $k = 0$, the problem is trivial. Indeed, if a sequence t that is distinct from s corresponds
to a permutation of the symbols of s, then s and t clearly share the same 1-mer profile. For
$k \geq 1$, an elegant answer can be found with the help of the de Bruijn subgraph DBG(s, k). \\ \newline
Clearly, by construction DBG(s, k) forms an Eulerian path $p_s$ , represented by its sequence
of vertices
\[
    s_{k,1}, s_{k,2}, \dots, s_{k, |s| -k+1}.
\]
% TODO nochmal klären und erklären was dieses s-k+1 jetzt ist und woher das kommt
Consequently DBG(s, k) is connected and balanced
and might contain even more than one Eulerian path. In fact, there is a one-to-one
correspondence between Eulerian paths in DBG(s, k) and sequences sharing the same
(k + 1)-mer profile. In other words, a sequence t that is distinct but has the same (k + 1)-
mer profile as s exists if and only if DBG(s, k) has another Eulerian path $p_t$ orresponding
to the sequence of (k + 1)-mers of t. Since $p_t$ uses the same edges as $p_s$ n a different
order, the path $p_t$ an only exist if DBG(s, k) contains a cycle. Note, however, that the
presence of a cycle is not a sufficient condition. Indeed, it is possible that DBG(s, k)
contains one or more cycles, but still admits only one Eulerian path. Some examples are
given in Figures \ref{fig:example-dbgs-2-1} and % todo zweite figure



\begin{exa} $\ $
\begin{figure}[H]
    \begin{subfigure}{0.33\textwidth}
        \begin{tikzpicture}[
            node distance=2cm,
            every node/.style={circle},
            label distance=2pt,
            >=Stealth
        ]

            \node (gat) at (-3.31,-3.94) {GAT};
            \node (att) at (-2.37,-1.37) {ATT};
            \node (tta) at (0,0) {TTA};
            \node (tac) at (2.37,-1.37) {TAC};
            \node (aca) at (3.31,-3.94) {ACA};

            \draw[->] (gat) to[bend left=20] (att);
            \draw[->] (att) to[bend left=20] (tta);

            \draw[->] (tta) to[bend left=10] (tac);
            \draw[->] (tac) to[bend left=10] (aca);

        \end{tikzpicture}
        \caption{$x = GATTACA, k = 3$ - Since this graph contains no cycle, it has only one Eulerian path. There can be no other word sharing the same 4-mer profile.} \label{fig:example-dbgs-2-1-a}
    \end{subfigure}
    \begin{subfigure}{0.33\textwidth}
        \begin{tikzpicture}[
            node distance=2cm,
            every node/.style={circle},
            label distance=2pt,
            >=Stealth
        ]

            \node (gat) at (-3.31,-3.94) {GAT};
            \node (att) at (-2.37,-1.37) {ATT};
            \node (tta) at (0,0) {TTA};
            \node (tac) at (2.37,-1.37) {TAC};
            \node (aca) at (3.31,-3.94) {ACA};
            \node (tat) at (-0.5,-1.5) {TAT};

            \draw[->] (gat) to[bend left=20] (att);
            \draw[->] (att) to[bend left=50] (tta);
            \draw[->] (att) -- (tta);

            \draw[->] (tta) to[bend left=10]  (tat);
            \draw[->] (tat) to[bend left=10]   (att);

            \draw[->] (tta) to[bend left=10] (tac);
            \draw[->] (tac) to[bend left=10] (aca);

        \end{tikzpicture}
        \caption{$x = GATTATTACA,  k = 3$ - Here the graph contains two cycles, but still only one Eulerian path. There is no other word sharing the same 4-mer profile.}
        \label{fig:example-dbgs-2-1-b}
    \end{subfigure}
    \begin{subfigure}{0.33\textwidth}
        \begin{tikzpicture}[
            node distance=2cm,
            every node/.style={circle},
            label distance=2pt,
            >=Stealth
        ]
            \node (gat) at (-3.31,-3.94) {GAT};
            \node (att) at (-2.37,-1.37) {ATT};
            \node (tta) at (0,0) {TTA};
            \node (tac) at (2.37,-1.37) {TAC};
            \node (aca) at (3.31,-3.94) {ACA};
            \node (tat) at (-0.5,-1.5) {TAT};

            \node (taa) at (-0.8,1.3) {TAA};
            \node (aat) at (-2.5,0.6) {AAT};

            \draw[->] (gat) to[bend left=20] (att);
            \draw[->] (att) to[bend left=50] (tta);

            \draw[->] (aat) to[bend right=50] (att);
            \draw[->] (tta) to[bend right=50] (taa);
            \draw[->] (taa) to[bend right=20] (aat);

            \draw[->] (att) -- (tta);
            \draw[->] (att) to[bend left=20] (tta);

            \draw[->] (tta) to[bend left=10]  (tat);
            \draw[->] (tat) to[bend left=10]   (att);

            \draw[->] (tta) to[bend left=10] (tac);
            \draw[->] (tac) to[bend left=10] (aca);

        \end{tikzpicture}
        \caption{$x = GATTATTAATTACA, k = 3$ - This graph contains several cycles and two distinct Eulerian paths. There is exactly one other word with the same 4-mer profile,
            \texttt{GATTAATTATTACA}.}
        \label{fig:example-dbgs-2-1-c}
    \end{subfigure}
    \caption{Examples of de Bruijn subgraphs for distinct sequences and k = 3.}\label{fig:example-dbgs-2-1}
\end{figure}
\end{exa}

\begin{exa}
    \textbf{Additional example from lecture}
    \begin{figure}[H]
        \begin{tikzpicture}[
            node distance=2cm,
            every node/.style={circle},
            label distance=2pt,
            >=Stealth
        ]

            \node (gat) at (-3.31,-3.94) {GAT};
            \node (att) at (-2.37,-1.37) {ATT};
            \node (tta) at (0,0) {TTA};
            \node (tac) at (2.37,-1.37) {TAC};
            \node (aca) at (3.31,-3.94) {ACA};
            \node (cag) at (1.5,-6) {CAG};
            \node (aga) at (-1.5,-6) {AGA};

            \draw[->] (gat) to[bend left=10] (att);
            \draw[->] (att) to[bend left=10] (tta);


            \draw[->] (tta) to[bend left=10] (tac);
            \draw[->] (tta) to[bend right=10] (tac);
            \draw[->] (tac) to[bend left=10] (aca);

            \draw[->] (aca) to[bend left=10] (cag);
            \draw[->] (cag) to[bend left=10] (aga);
            \draw[->] (aga) to[bend left=10] (gat);

        \end{tikzpicture}
        \caption{It could also be on big circle.}\label{fig:figure2}
    \end{figure}
\end{exa}

\begin{kom}
    Reminder that the edges are basically this:
    \begin{figure}[H]
        \begin{tikzpicture}[
            node distance=2cm,
            every node/.style={circle},
            label distance=2pt,
            >=Stealth
        ]

            \node (gat) at (-3.31,-3.94) {GAT};
            \node (att) at (-2.37,-1.37) {ATT};


            \draw[->, WildStrawberry] (gat) to["GAAT"] (att);

        \end{tikzpicture}\label{fig:figure3}
    \end{figure}
\end{kom}

\begin{theo}
    Theorem 2.1 auf deutsch: \# worte mit gleichen (k+1)-mer Profil wie s = \# Euler-Pfade in DBG(s,k)
\end{theo}

\begin{kom} \textbf{To figure \ref{fig:example-dbgs-2-1}}
    man könnte bei fig \ref{fig:example-dbgs-2-1-a} denken hey bei a kann es nur einen EP geben, weil es kein Kreis
    ist, und bei \ref{fig:example-dbgs-2-1-b}
    muss es dann
    mehrere geben, weil da eine Schleife drin ist, aber NEIN der Pfad ist die Reihenfolge der Knoten und nicht der
    Kanten, also gibt es nur einen. % todo (visualisieren)
    Man kann nicht unterscheiden, ob man zuerst die obere oder untere
    Kante der Schleife zuerst gegangen ist.
    Gibt also nur eine sequenz mit em k+1 meer profil.
    Denn die Sequenz wird
    ja aus den Knoten gemacht, nicht den Kanten \ldots sozusagen.
\end{kom}

\begin{kom}
    Der Begriff dbg ist sehr dehnbar, es ist in der Literatur nicht immer das gleiche damit gemeint.
\end{kom}


\section{Variants of de Bruijn Graphs}\label{sec:variants-of-de-bruijn-graphs}

\defin{origdb}

\begin{exa}
    zu origdb
\end{exa}


\begin{exa}
    zu dbsg
\end{exa}

\begin{prob}
    mit dem bi dings, anhang von bsp auf notiz
\end{prob}

\defin{bidbsg}

\begin{exa}
    zu bidbsg
\end{exa}


\defin{comdbsg}

\begin{exa}
    zu comdbsg
\end{exa}

\defin{coldbsg}

\begin{exa}
    zu coldbsg
\end{exa}

\begin{prob}
    log der alphabetgröße zur basis 2 ist wie viele bits man braucht zum codieren
    k*2 bits für dbg, aber mit Farben muss man ja noch bits dran hängen, das ist groß und noch ungelöst, denn es ist
    dann k*2 + \# datensätze
\end{prob}

\begin{exa}
    bsp zu bits: acg codiert als 000110 wenn

    \begin{tabular}{lll}
        A & 0 & 0 \\
        C & 0 & 1 \\
        G & 1 & 0 \\
        T & 1 & 1
    \end{tabular}

\end{exa}


\section{Implementation of de Bruijn Graphs: Sets of k-mers}\label{sec:implementation-of-de-bruijn-graphs:-sets-of-k-mers}


